GPU-friendly Stroke Expansion

### 2024-04-22 paper.tex
Input encoding section text

\documentclass[sigconf, authordraft]{acmart}
\usepackage[linesnumbered]{algorithm2e}
\begin{document}
\title{GPU-friendly Stroke Expansion}
\author{Raph Levien}
\author{Arman Uguray}
\affiliation{
    \institution{Google}
    \city{San Francisco}
    \state{CA}
    \country{USA}
}

@@

\subsection{Pipeline Design}

@@

% TODO: caps/joins figure highlighting mapping of segments, joins, and caps to threads, encoding. Make sure to show the straight-line inner join.
% TODO: visualize per-segment flattened output for a fill and for a stroke from the perspective of one thread.
% TODO: finish this \begin{algorithm} snippet
\begin{algorithm}
  $tid \gets$ global invocation ID\;
  $(style, segment) \gets readScene(tid)$\;
  \eIf{$style$ is $fill$}{
      flatten source curve
  }{\If{you're sad}{
      just keep going
  }}
\caption{The control flow of our compute shader}
\end{algorithm}

@@

The joins between adjacent segments need to form a continuous outline when combined with the segments. This requires that a thread have access to the tangent vector at the end of its assigned segment \emph{and} the start tangent of the next adjacent segment. We require that the input segments within a subpath are stored in order which simplifies the access to the adjacent segment down to a buffer read at the next array offset. Notably, this scheme does not cooperation across threads. While we take care to output the \emph{outer} join segment (where the offset curves meet at a reflex angle) we output the inner join as a line segment.

@@

% TODO: not sure how much to emphasize the path tag encoding. The input processing monoids benefit the vello scene API (fragments) the TagMonoid could generally be computed on the CPU (with minimal offset tracking) It's not that relevant to the stroke processing. The main bit to mention is that it's a struct-of-arrays.
% TODO: talk about sanitization of input: closepaths are handled ahead of time, stroke cap marker.
% TODO: discuss style encoding in greater detail

One contribution of this paper is a highly efficient encoding for paths to be rendered. Goals include conciseness, low cost to encode on the CPU side, and fully parallel processing by the GPU. There are two major aspects to the encoding: a variable size encoding of paths, suitable for both filling and stroking, and an extension to stroking. The input is separated into a \emph{tag stream}, with one byte per path element, and another stream for the coordinate data. The index into coordinate data is computed by a prefix sum from a size value derived from the tag byte.

\subsection{Path encoding}

The tag byte is a number of bit fields, packed into a byte.

TODO: make this into a diagram.

\begin{itemize}
    \item 0-1: coordinate count
    \item 2: subpath end bit
    \item 3: 16/32 bit coordinates
    \item 4: path end bit
    \item 5: transform bit
    \item 6: line style bit
\end{itemize}

For most path segments, the coordinate count indicates the type of path segment: 1 for lines, 2 for quadratic Beziers, and 3 for cubic Beziers. The subpath bit is set on the last segment of a subpath, and the path bit is additionally set on the last bit of a path. The latter is not directly interpreted by the path flattening code, but is forwarded to later rendering stages, for example, the path id can be used to select a color or other paint style.

The size is readily inferred from the tag byte: the number of coordinate pairs is the coordinate count plus one if the subpath end bit is set. The nuber of bytes per coordinate pair is 4 for 16 bit coordinates or 8 for 32 bit coordinates.

It is relatively straightforward to convert the standard moveto/lineto/curveto path representation into this encoding. Subpath boundaries are moved from the beginning of a subpath (moveto) to the end. A separate segment must be generated for closepath if the ending point does not coincide with the start point.

The handling of the subpath bit allows for overlap in coordinates, so that, except at subpath boundaries, the first coordinate pair of each path segment overlaps with the last coordinate pair of the previous one. Subpath handling is all computed in the prefix sum; no special handling is needed on the GPU side.

\subsection{Algorithm}

@@

\subsection{Limitations}

% TODO: memory usage discussion: the impact of lines and how arcs might help. or work graphs, or maybe mesh shaders. Limitation: memory must be allocated up-front. bump estimation, violates the no pre-curve-evaluation criterium but simpler techniques (like Wang's) exist and we use it but way over-estimates. This is a limitation of the GPU execution model. Explain this trade-off.
  % -> TODO: Memory usage graphs for input scene buffers and LineSoup.

@@

\section{Discussion}

>>>>

\documentclass[sigconf, authordraft]{acmart}
\usepackage[linesnumbered]{algorithm2e}
\usepackage[format=plain, labelfont={bf}, textfont=normal]{caption}
\usepackage{booktabs,xcolor}
\begin{document}
\title{GPU-friendly Stroke Expansion}
\author{Raph Levien}
\author{Arman Uguray}
\affiliation{
    \institution{Google}
    \city{San Francisco}
    \state{CA}
    \country{USA}
}

@@

\subsection{Pipeline Design} \label{subsec:pipeline-design}

@@

% TODO: caps/joins figure highlighting mapping of segments, joins, and caps to threads, encoding. Make sure to show the straight-line inner join.
% TODO: visualize per-segment flattened output for a fill and for a stroke from the perspective of one thread.
% TODO: finish this \begin{algorithm} snippet
\begin{algorithm}
  $tid \gets$ global invocation ID\;
  $(style, segment) \gets readScene(tid)$\;
  \eIf{$style$ is $fill$}{
      output source curve\;
  }{
      \tcp{Handle stroked segment}
      \eIf{$segment$ is \emph{stroke cap marker}}{
          \eIf{path is open}{output start cap\;}{output nothing\;}
      }{
          output both offset curves\;
          $neighbor \gets readScene(tid + 1)$\;
          \eIf{$neighbor$ is not \emph{stroke cap marker} OR path is closed}{
              output join\;
          }{
              output end cap\;
          }
      }
  }
\caption{The control flow the compute shader}
\label{alg:shader-flow}
\end{algorithm}

@@

The joins between adjacent segments need to form a continuous outline when combined with the segments. This requires that a thread have access to the tangent vector at the end of its assigned segment \emph{and} the start tangent of the next adjacent segment. We require that the input segments within a subpath are stored in order which simplifies the access to the adjacent segment down to a buffer read at the next array offset. While we take care to output the \emph{outer} join segment (where the offset curves meet at a reflex angle) according to the join style, we output the inner join simply as a line segment.

@@

The input to our pipeline consists of a sequence of paths, each with an associated 2D affine transformation matrix and a \emph{style} data structure containing information about the stroke style or fill rule. We encode paths, transforms, and styles as parallel input streams. Unlike a conventional structural-of-arrays arrangement, we permit a one-to-many mapping across elements between certain streams: each transform and style entry can apply to one or more path entries with an array index greater than or equal to transform or style index.

Paths get encoded as two parallel streams: a \emph{path tag} stream containing an element for every segment in every subpath, and a \emph{path data} stream of point coordinates. Each segment/verb contributes a variable number of points: 1 for \emph{moveto}/\emph{lineto}, 2 for quadratic Béziers, and 3 for cubic Béziers. This variable length stream encoding allows for a highly compact representation of SVG-type content in memory. A GPU thread assigned to a path segment must be able to reference the corresponding element in each stream in order to obtain the right curve coordinates, transform, and style information. This is done by computing a set of stream offsets for every element in the path tag stream.

% TODO: Figure: streams and tagmonoid relationship, plus sizes of the elements, plus monoid bits

We compute the offsets on the GPU in a separate compute shader that runs prior to the flattening stage. A path tag consists of 8 bits that store \emph{stream offset increments}, such that the inclusive prefix sums of all increments for every element in the tag stream yield the stream offsets. We call this structure the \emph{tag monoid} and have implemented the computation as a general parallel prefix scan.[TODO: citation?]

\begin{table}
    \begin{tabular}{|c|l|}
    \hline
        \textbf{Path Tag Bits} & \multicolumn{1}{|c|}{\textbf{Monoid Fields}} \\
    \hline
        0-1 & coordinate count \\
        2 & subpath end bit \\
        3 & 16/32 bit coordinates \\
        4 & path bit \\
        5 & transform bit \\
        6 & style bit \\
    \hline
    \end{tabular}
    \caption{
        \emph{Coordinate count} indicates the type of path segment: 1 for lines, 2 for quadratic Béziers, 3 for cubic Béziers. The \emph{subpath end bit} is set on the last segment of a subpath and the \emph{path bit} is additionally set on the last tag of a path. The number of curve control points is given by \emph{coordinate count}, plus $1$ if the subpath end bit is set. The number of bytes per coordinate pair is 4 for 16 bit or 8 for 32 bit coordinates.
    }
    \label{table:pathtag}
\end{table}

Table~\ref{table:pathtag} shows the individual fields of the 8-bit path tag. The offset increments for the transform and style streams can be stored in a single bit that is set to $1$ only if a path tag marks the beginning of a new transform and/or style entry. We also encode an additional \emph{path bit} in the final segment of all paths, which allows us to access additional streams (such as a fill color or gradient parameters) using a per-path offset in later rendering stages.

The handling of the subpath bit allows for overlap in coordinates, so that except at subpath boundaries, the first coordinate pair of each segment overlaps with the last coordinate pair of the previous one. During encoding, subpath boundaries are moved from the beginning of a subpath (moveto) to the end. An additional line segment is inserted into the tag stream for fills (closepath) if the final point does not coincide with the start point. The final segment of every subpath has the \emph{subpath end bit} set to $1$. For strokes, the subpath end segment represents the \emph{stroke cap marker} defined in Section \ref{subsec:pipeline-design}. This special segment is assigned a coordinate count of $2$, and the corresponding two points in the path data stream encode the tangent vector of the subpath's initial segment, used to correctly draw the start cap or the connecting join in a closed contour.

\subsection{Flattening Algorithm}

@@

% TODO: memory usage discussion: the impact of lines and how arcs might help. or work graphs, or maybe mesh shaders. Limitation: memory must be allocated up-front. bump estimation, violates the no pre-curve-evaluation criterium but simpler techniques (like Wang's) exist and we use it but way over-estimates. This is a limitation of the GPU execution model. Explain this trade-off.
  % -> TODO: Memory usage graphs for input scene buffers and LineSoup.
% TODO: pathological cases: our implementation doesn't handle culling.

@@

\section{Discussion}
\subsection{Limitations}
